---
title: "R: using NSE with select helpers and default arguments"
draft: false
categories: ["R"]
tags: ["R", "snippets", "tidyverse", "NSE"]
date: "2018-10-11"
output:
   blogdown::html_page:
      default
---
   
   
```{r setup, include=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))

knitr::opts_chunk$set(collapse = TRUE)
```

## Introduction

Once you adopt the {{<cran tidyverse>}} philosophy into your R code, at a certain time you will start writing functions which invoke [dplyr's non-standard evaluation (NSE)](https://dplyr.tidyverse.org/articles/programming.html) mechanisms.

An example is the following. Consider this dataset:

```{r, eval=FALSE}
library(tidyverse)

storms
```
```{r, echo=FALSE}
storms %>% 
   head(10) %>% 
   kable() %>% 
   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11, full_width = TRUE)
```

Suppose that you want to select all columns related to time (i.e. `year`, `month`, `day`, `hour`).      
Using the standard {{<cran dplyr>}} tools, this is done using `dplyr::select`:

```{r}
storms_ymdh <- storms %>% select(year, month, day, hour)
storms_ymd <- storms %>% select(year, month, day)
storms_ym <- storms %>% select(year, month)
storms_y <- storms %>% select(year)
```

This can be incapsulated into a function, as usual:
```{r}
fun_select <- function(df) {
   df %>% select(year, month, day, hour)
}

identical(storms_ymdh, storms %>% fun_select())
```

`select` provides several helpers to indicate which variables are kept.    
An example is `one_of()`:

```{r}

identical(storms_ymdh, 
          storms %>% select(one_of(c('year', 'month', 'day', 'hour'))) )

fun_select_2 <- function(df) {
   df %>% select(one_of(c('year', 'month', 'day', 'hour')))
}

identical(storms_ymdh, storms %>% fun_select_2())
```

### Adding arguments

Now, we want to generalize the mechanism, and allow the user to decide which variables are selected.   

The easiest adaptation does not use NSE at all, we just pass the vector of columns to select:

```{r}
fun_select_3 <- function(df, cols) {
   df %>% select(one_of(cols))
}

identical(storms_ymdh, storms %>% fun_select_3(c('year', 'month', 'day', 'hour')))
identical(storms_ymd, storms %>% fun_select_3(c('year', 'month', 'day')))
```

### Enter the NSE

By reading up on NSE, we can make `fun_select_3` behave as `select` behaves, without explicitly quoting the argument:

```{r}
fun_select_NSE <- function(df, ...) {
   dots <- enquos(...)
   df %>% select(!!!dots)
}

identical(storms_ymdh, storms %>% fun_select_NSE(year, month, day, hour))
identical(storms_ymd, storms %>% fun_select_NSE(year, month, day))
```
It also works with all `select` helpers:
```{r}
# one_of
identical(storms_ymd, storms %>% fun_select_NSE(one_of(c('year', 'month', 'day'))) )
identical(storms_ymd, storms %>% select(one_of(c('year', 'month', 'day'))) )

# Column range
identical(storms_ymdh, storms %>% fun_select_NSE(year:hour))
```

#### Why it works

The three-dot parameter `...` captures all arguments[^1] of `fun_select_NSE`. 
Then `enquos` quotes the list, essentially blocking evaluation until an unquoting operation appears.   
In this case, the evaluation continues when the `!!!` is encountered: that operator inserts the names in place of the quoted variables.

[^1]: *parameter*: what is declared, *argument*: what is actually passed.

Also, `!!!` performs a *splicing* operation: arguments are added as to the surrounding function, separated by commas. (It is similar to Python's unpacking operator `*`). If we were to accept a single argument, instead, we could have just used `enquo` and `!!`.

Notice that the mechanism works for any kind of expression, like `one_of(c('year', 'month'))`.   
First, `...` are set to be equal to the argument, but it is not evaluated because of `enquos` effect.
`!!!`, then, writes back the user's expression as `select`'s argument.

To know more about `tidyverse`'s `quo`/`enquo`/quotation/quasiquotation/`!!`, please read [this wonderful post](https://milesmcbain.xyz/the-roots-of-quotation/)!   
More R-ready material is available in Hadley Wickham's [Advanced R](https://adv-r.hadley.nz/quasiquotation.html), and the [quasiquotation page](https://www.rdocumentation.org/packages/rlang/versions/0.2.1/topics/quasiquotation) in {{<cran rlang>}}-package documentation.


### Modifying default parameters with NSE

However, we would like to provide a default parameter which performs some selection.    
E.g. select only `year` if no argument is passed.

It is very easy to do for the non-NSE functions, if you stick to strings:
```{r}
fun_select_3_default <- function(df, cols_default = 'year') {
   df %>% select(one_of(cols_default))
}

identical(storms_ymdh, storms %>% fun_select_3_default(c('year', 'month', 'day', 'hour')) )
identical(storms_y, storms %>% fun_select_3_default() )
```

If you want to pass more complicated expressions, however, the NSE version is a little trickier:

```{r}
fun_select_NSE_default <- function(df, ...) {
      
   dots <- enquos(...)

   if (length(dots) != 0) {
      # What is passed if dots are not empty
      select_var_true <- dots
   } else {
      # Default value: year
      select_var_true <- quo(year)
   }
   
   df %>% select(!!!select_var_true)
}
identical(storms_ym, storms %>% fun_select_NSE_default(year, month))
identical(storms_y, storms %>% fun_select_NSE_default())
```

We can also use `rlang::quo_is_missing`, which returns `TRUE` if the quosure is empty (in this case, when nothing lies in the place of `...`):

```{r}
fun_select_NSE_default_2 <- function(df, ...) {
      
   dots <- enquos(...)

   if (all(purrr::map_lgl(dots, rlang::quo_is_missing))) {
      # Default value: year
      select_var_true <- quo(year)
   } else {
      # What is passed if dots are not empty
      select_var_true <- dots
   }
   
   df %>% select(!!!select_var_true)
}
identical(storms_ym, storms %>% fun_select_NSE_default_2(year, month))
identical(storms_y, storms %>% fun_select_NSE_default_2())
```

Note that `purrr::map` is necessary, since we are dealing with multiple comma-separated parameters (captured by the dots).
`rlang::quo_is_missing` works only with a single quosure, `dots` is a list of quosures!

We can also set default parameters which are more complex than a single variable name, e.g., a select helper:

```{r}
fun_select_NSE_default_3 <- function(df, ...) {
      
   dots <- enquos(...)

   if (all(map_lgl(dots, rlang::quo_is_missing))) {
      # Default value: year, month
      select_var_true <- quo(one_of(c('year', 'month')))
   } else {
      # What is passed if dots are not empty
      select_var_true <- dots
   }
   df %>% select(!!!select_var_true)
}
identical(storms_ym, storms %>% fun_select_NSE_default_3())
```

Thanks for reading!