---
title: "Viewing scraped HTML sessions from rvest and friends"
draft: true
categories: ["coding"]
tags: ["R", "rvest"]
date: "2018-01-17"
editor_options: 
  chunk_output_type: console
---


Hadley Wickham's excellent `rvest` R package is an extremely easy way to scrape any HTML page.    
It also provides limited utilities to do the scraping interactively: 

- `html_session`, which actually starts a browsing session in R interpreter,
- `back`, `follow_link`, `jump_to` implement page navigation using history or link selectors (CSS/XPath)
- `html_nodes` accept `session` objects and return HTML nodes in `rvest`-friendly format

However, there is no easy way to see the code which is actually parsed.  
Since `rvest` uses `httr` to make HTML requests, returned code can be very different from what is seen on a browser.

Similar issues exist even if you use other R packages such as `httr` (which, basically, deals only with the HTTP requests), `curl` or `RCurl` (same, but the hard way!).

In this post, we provide some quick functions to access the HTML code and view it in a browser. We will also learn how to exploit the S3 object system in R to help us!

### Using `rvest`

Suppose that we have just scraped a page:
```{r}
library(xml2)
library(rvest)

url <- 'https://nghttp2.org/httpbin/'
s <- rvest::html_session(url)
class(s)
```

`s` holds the scraped page (it is a ` `r class(s)` `-object), we can access the page contents using `xml2::read_html()`:

```{r}
s_tree <- xml2::read_html(s)
class(s_tree)
s_tree
```

We can see that `s_tree` has class `xml2::xml_document`, but the code is still hidden. The function `as.character` helps:

```{r}
as.character(s_tree)
```

Here it is!
Now we want to view it on a browser. 

R `base` package contains the function `browseURL()`, which however opens the browser to an URL or a file. So we have to store it on disk. 

Let's create a function which does all of this: getting a `session` object, extracting the HTML code, casting to character, writing it to disk and taking care of cleaning the output.

```{r, eval=FALSE}
view_rvest <- function(s) {

   # Cast the session to character
   stopifnot(class(s) == 'session')
   s_tree <- xml2::read_html(s)
   s_code <- as.character(s_tree)
   
   # Make a temporary file, fill it with text
   temp_file <- tempfile(fileext = '.html')

   f <- file(temp_file, open = 'w')
   write(s_code, f)
   close(f)
   
   browseURL(temp_file)

   # Wait a while, then delete it
   Sys.sleep(3)
   unlink(temp_file)
}

view_rvest(s)
```

And voilà!


### Using `httr`

Suppose, now, that we want to do the same with an `httr` request.   
E.g., the code from the previous page can be retrieved by:

```{r}
req <- httr::GET(url)
httr::content(req, as = 'text')
```

where `req` is a ` `r class(req)` `-object.

The resulting do-it-all function would be:

```{r, eval = FALSE}
view_httr <- function(req) {

   # Cast the session to character
   stopifnot(class(req) == 'response')
   s_code <- httr::content(req, as = 'text')
   
   # Make a temporary file, fill it with text
   temp_file <- tempfile(fileext = '.html')

   f <- file(temp_file, open = 'w')
   write(s_code, f)
   close(f)
   
   browseURL(temp_file)

   # Wait a while, then delete it
   Sys.sleep(3)
   unlink(temp_file)
}

view_httr(req)
```

And voilà! (2)

### Generalizing with S3 objects

So far, we wrote two functions:

- `view_rvest`, operating on `text` objects
- `view_httr`, operating on `request` objects

We would like to have a single function which elegantly deals with both cases, and allows for easy extensions.  

Either we write a wrapper, and distinguish between the two with an `if`, or we use the [S3 object](https://adv-r.hadley.nz/s3.html) system in R.

It is actually very simple and hackish, but it works well!

In short, we write a **generic** method, which then calls (**dispatches**) the appropriate method for the object type.   
The generic method will be seen by the user, the following methods will be, in general, hidden.

```{r}
view_html <- function(x) {
   UseMethod('view_html', x)
}
```

What it does when called, is to grab the class of `x`, and call the method named as `view_html.class` (where, of course, class is `class(x)`).   
So, we just need to rename:

- `view_httr` to `view_html.response` (as `x` is a `response`, from `httr`) 
- `view_rvest` to `view_html.session` (as `x` is a `session`, from `rvest`)

To be more compact, we can also introduce the `view_html.character` method which writes character vectors to disk and opens a browser.

Finally, here are the resulting methods:

```{r, eval = FALSE}
view_html <- function(x) {
   UseMethod('view_html', x)
}
view_html.session <- function(s) {
   s_tree <- xml2::read_html(s)
   s_code <- as.character(s_tree)
   
   view_html(s_code)
}
view_html.response <- function(resp) {
   s_code <- httr::content(resp, as = 'text')
   
   view_html(s_code)
}

view_html.character <- function(text) {
   temp_file <- tempfile(fileext = '.html')
   
   print(paste('Created:', temp_file))
   
   f <- file(temp_file, open = 'w')
   write(text, f)
   close(f)
   
   browseURL(temp_file)
   Sys.sleep(3)
   unlink(temp_file)
}

# The calls (hidden from the user!):
view_html(s)
view_html(req)
view_html('test')
```

Thank you for reading!