---
title: "Controlling long outputs with knitr hooks"
draft: false
categories: ["blog"]
tags: ["R", "blogdown", "knitr", "hooks", "snippets"]
date: "2018-05-30"
---


```{r setup, include=FALSE}
library(magrittr)
knitr::opts_chunk$set(collapse = TRUE)

# source('set_hooks.R', local = FALSE)
```

## Introduction

This blog is written in a combination of Markdown and R Markdown (using the R package `r blogdown::shortcode('cran','blogdown')`).

As such, some R Markdown posts contain code chunks who generate very lengthy and verbose output.   
An example is this one, which generates a sequence of strings separated by newlines:
```{r long-lines}
n <- 10
m <- 10
matrix(rnorm(m*n), m, n)
```

and this one, which generates a single long string:
```{r long-string}
long_string <- paste(rep(letters, 30), collapse = '')
long_string
```

## Enter: knitr hooks

As detailed by the [knitr documentation](https://yihui.name/knitr/hooks/), `knitr` provides a set of functions to transparently modify a chunk, without changing the inner contents, but specifying chunk options. Such functions are called **hooks**.

Knitr hooks can be of two types: chunk hooks, output hooks and option hooks.   
To our purpose, we are interested in **output hooks**.   

These are functions which, in general, take two parameters: the chunk output, and the chunk options (e.g. `fig.width`).

From the [documentation](https://yihui.name/knitr/hooks/#output-hooks), an output hook can deal with 8 different types of outputs:

- *source*: the source code
- *output*: ordinary R output (i.e., what would have been printed in an R terminal) except warnings, messages and errors
- *warning*: warnings from warning()
- *message*: messages from message()
- *error*: errors from stop() (applies to errors in both code chunks and inline R code)
- *plot*: graphics output
- *inline*: output of inline R code
- *chunk*: all the output of a chunk (i.e., those produced by the previous hooks)
- *document*: the output of the whole document (is base::identity by default)

Again, we are interested in output hooks for the output, as we want to preserve source code, errors, warnings and messages.

### Line truncation

This is directly taken from knitr examples: [`052-suppress-output.Rmd`](https://github.com/yihui/knitr-examples/blob/master/052-suppress-output.Rmd)

First, e.g. in the setup chunk, we retrieve the default output hook:
```{r}
# the default output hook
hook_output_default <- knitr::knit_hooks$get('output')
```

We define a function which truncates the output up to `n` lines. `x` is the chunk output, as a character vector. If `n` is missing, the output passes through, unaffected.

```{r}
truncate_to_lines <- function(x, n) {
   if (!is.null(n)) {
      x = unlist(stringr::str_split(x, '\n'))
      if (length(x) > n) {
         # truncate the output
         x = c(head(x, n), '...\n')
      }
      x = paste(x, collapse = '\n') # paste first n lines together
   }
   x
}
```

Finally, we can overwrite the default knitr hook, and truncate if option `max.lines` is specified:
```{r}
knitr::knit_hooks$set(output = function(x, options) {
   max.lines <- options$max.lines
   x <- truncate_to_lines(x, max.lines)

   hook_output_default(x, options)
})
```

Let's see the first chunk, now truncated to 3 lines. We set `max.lines=3` in the chunk options:

~~~~md
`​​``{​r, max.lines = 3}
n <- 10
m <- 10
matrix(rnorm(m*n), m, n)
`​``
~~~~

and here is the output:

```{r long-lines, max.lines = 3}
```


### Character truncation

This time we need to define a function which truncates up to a specified amount of characters, if `n_chars` is specified:
```{r}
truncate_to_chars <- function(x, n_chars) {
   if (!is.null(n_chars)) {
      if (min(nchar(x), n_chars) < nchar(x)) {
         x <- substr(x, 1, min(nchar(x), n_chars))
         x <- paste(x, ' ...\n')
      }
   }
   x
}
```

Then, we can add the truncation function to the output hook:
```{r}
knitr::knit_hooks$set(output = function(x, options) {
    max.lines <- options$max.lines
    x <- truncate_to_lines(x, max.lines)

    # Here is our new code
    max.chars <- options$max.chars
    x <- truncate_to_chars(x, max.chars)

    hook_output_default(x, options)
})
```

And here is the second chunk, truncated to 20 chars (setting `max.chars = 20` in the chunk options):
```{r long-string, max.chars = 20}
```
Notice that the code is not part of the character count (it is an output hook on the *output*!), but the prompt and the whitespaces are (up to `"..."`).

----

And, of course, both truncations can be combined.   
E.g., let's see the first chunk to 3 lines and 160 characters:
```{r long-lines, max.lines=3, max.chars=160}
```

Thanks for reading!